\Huge \textbf{Species sympatry shapes brain size evolution in Primates}

### 2021-08-11 Article/Article.Rmd
Rerun pgls with correct frugivory threshold and shortened manuscript

---
title: \Huge \textbf{Species sympatry constrains brain size evolution in Primates}
# date: "`r format(Sys.time(), '%B %d, %Y')`"

#I follow: https://stackoverflow.com/questions/52918716/authors-and-affiliations-in-the-yaml-of-rmarkdown for author display
author: 
  - name: \textit{Benjamin Robira}
    email: benjamin.robira@normalesup.org
    institute: [CEFE, MH]
    correspondence: true
  - name: \textit{Benoit Perez-Lamarque}
    email: benoit.perez@ens.psl.eu
    institute: [ENS, MNHN]
    correspondence: true
institute:
  - CEFE: Centre d’Écologie Fonctionnelle et Évolutive, Université de Montpellier & CNRS, Montpellier, France.
  - MH: Anthropologie et Ethnobiologie, Centre National de la Recherche Scientifique/Muséum National d'Histoire Naturelle, University Paris Diderot, Sorbonne Paris Cité, Musée de l'Homme, Paris, France
  - ENS: Institut de Biologie de l’École Normale Supérieure (IBENS), École Normale Supérieure, CNRS, INSERM, Université PSL, Paris, France
  - MNHN: Institut de Systématique, Évolution, Biodiversité (ISYEB), Muséum National d’Histoire Naturelle, CNRS, Sorbonne Université, EPHE, Université des Antilles, Paris, France

@@

bibliography: bibliographyarticlepackage.bib
csl: science.csl
#biblio-style: apa
always_allow_html: true
link-citations: yes

@@

header-includes:
# - \usepackage[nolists, nomarkers,tablesfirst]{endfloat} # For figures and tables at end
  - \usepackage{lineno} # For line numbering
  - \linenumbers # For line numbering
  - \usepackage{setspace}\doublespacing
  - \setlength{\parskip}{0em} #to remove line gaps
  - \DeclareUnicodeCharacter{2212}{-}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \newcommand{\beginsupplement}{
      \setcounter{table}{0}  
      \renewcommand{\thetable}{S\arabic{table}} 
      \setcounter{figure}{0} 
      \renewcommand{\thefigure}{S\arabic{figure}}
    }
---

@@

**Abstract:** The diversity in animal cognition raises the question of its underlying evolutionary drivers. Selection upon more advanced cognitive abilities can stem from interactions of individuals with conspecifics within the social unit (*Social Intelligence Hypothesis*), among generations (*Cultural Intelligence Hypothesis*), between social units (*Napoleonic Intelligence Hypothesis*), or with the rest of their environment (*Ecological Intelligence Hypothesis*). These hypotheses were limited to within-species scenarios. Yet, one species rarely occupies an area alone: Space is a place shared between many species that can interact directly or indirectly. For species occupying a same dietary niche, interspecific competition could induce a cognitive evolutionary arms race, so as to outperform competitors. Furthermore, all species act upon the spatio-temporal distribution of resources, hence contribute to increasing the uncertainty for a third species to infer resource location and availability. This would ultimately shape the cognitive machinery involved to allow sufficient harvesting. In addition, sympatric species can leave traces of passage and of resource presence that can be cognitively processed by a third species to find it more efficiently. As such, to test whether species co-occurrence shaped current patterns of cognition, we used primates as a study example and retraced the evolutionary history of multiple species brain areas involved in foraging activities or not while considering competitive or non-competitive evolutionary scenarios. We found that the evolution of the relative size of areas involved in foraging-related information processing and/or retention, as well as in areas related to processing social information, are better described by models accounting for species co-occurrence within dietary guilds. More precisely for these brain areas, species co-occurrence was associated to a decrease, and never to an increase, of their relative size. Coherently with the observed wide variability in sympatry rate and intensity, the degree of encephalisation was unrelated to the evolutionary success of a lineage (i.e. its diversification rate). Overall, this comparative study suggests that species co-occurrence stands as a brake to positive selection towards larger cognitive abilities, yet leaves open the question of the underlying ecological mechanisms at play.

<!-- subject to an arms race among species (*Red Queen Intelligence Hypothesis*).  -->

**Keywords:** "Brain size - Cognition - Cooperation - Competition - Intelligence evolution - Primates - Species co-occurrence"

@@

<!-- TC:ignore -->
**Word Count:** `r words$num_words` \newline
<!-- Character Count: `r words$num_char` -->
<!-- TC:endignore -->

@@

On the road to brain size evolution, generally considered as an equivalent of cognition evolution, mysteries are plenty [@dunbar2017there]. It remains puzzling why humans have a brain that is that large, relatively to body size, compared to other animals, or why primate brain architecture, in general, is much more complex than observed in other taxa. Overall, the evolution of the brain is the consequence of constraints [e.g. energetic limitations, @navarrete2011energetics], but also of socio-ecological drivers promoting cognitive abilities [@gonzalez2018inference].

|   To grasp the rationale of current hypotheses that aim to describe the evolution of cognition [@van2006some; @dunbar2017there], consider a primate individual foraging for food. Primates are pivotal species to study the evolution of cognition given the complexity of their socio-ecological environment and the inevitable implication for retracing human evolutionary history [@byrne2000evolution]. The chosen focal is not looking for any type of food but specific one, as for instance, most primate species often look for fruits. Finding fruits for a primate is not an easy task. In fact, fruits are the archetype of a hard-to-find resource likely to promote cognitive development because fruit trees are rare, dispersed, and do not produce constantly along the year nor between years, albeit their fruiting period remains nonetheless predictable [@janmaat2016spatio]. Additionally, they are energy-rich, allowing expansion of costly tissues such as brain tissues [@clutton1980primates]. Overall, primates thus show remarkable abilities to navigate precisely and target fruit trees likely to yield ripe resource despite they are out of their sensory range [@trapanese2019review], in part because of advanced spatio-temporal information retention [e.g. spatial knowledge: @normand2009forest; @robiraroute, temporal knowledge: @janmaat2006evidence; @janmaat2013chimpanzees; @janmaat2013tai].

|   When the resource is out of sight, moving at random could mean travelling unnecessary long distances to finally reach a potentially void tree. Hence, for the focal, being capable of processing immediate environmental cues and clues to know where specifically to find fruit trees [@dall2005information; @grove2013evolution] and whether it is probable that these trees currently yield fruits [@dall2005information; @robirainreview; @janmaat2016spatio] would be a considerable advantage. Having an appropriate cognitive machinery may avoid costly detours and time recursions to food patches adequately. This simple picture draws the basis of the *Ecological Intelligence Hypothesis* [@clutton1980primates; @milton1981distribution; @rosati2017foraging] which stipulates that cognition was selected as a way to buffer the spatio-temporal complexity of the environment (e.g. @van2012large).

|   In its quest for food, however, the focal might not be alone: For instance, primates often form social groups [@kappeler2002evolution] in which group members spend considerable time together, have established bounds and constantly share information. Thus, they might altogether look for, and process, cues and clues leading to food sources. To do so, being able to process signals emitted by other maneuvers would be an advantage, as well as reading through their mind [@devaine2017reading] in order to plan ahead for collective actions or even outsmart them [@byrne1994machiavellian]. This therefore means plenty of additional information to process, and constitutes the socle of the *Social Intelligence Hypothesis* [@dunbar2017there].

|   In the focal group, however, all individuals are not equally knowledgeable: Perhaps because some have more experience than others, such as the elderly ones would be more experienced than the youths, with the latter thus learning from the former [@laland2004social]. For instance, young individuals may progressively learn how to process a specific tool to access a hidden comestible part as for nuts [@boesch1994nut] by observing and reproducing, or possibly being teached (learning ways detailed in @estienne2019acquisition). In other words, there is vertical transmission that passes by over generations and knowledge thereby accumulates: This is the *Cultural Intelligence Hypothesis* [@whiten2007evolution; @reader2002social], also known as the *Vygotskian Intelligence Hypothesis* in humans [@herrmann2007humans; @tomasello2009cultural; @wilson1991molecular; @van2011social].

|   Finally, while the group peacefully forages, it might sense the past or current presence of another group and anticipate its move, such as wild baboons do when prioritizing food likely to be depleted by other troops [@noser2010wild]. The neighboring groups might indeed be competitors for food, mates or any essential resource, and as much as it is advantageous to be able to read through the environment and intra-group companions, it might be advantageous to be able to decode information relative to other groups’ presence and attribute (e.g. group size), a view brought by the *Napoleonic Intelligence Hypothesis* [@ashton2020interactions].

|   Yet, in this overly simplistic picture, we moved from a unique individual to a group of individuals and then to multiple groups. Never was it question of multiple species. Yet, space is a place shared between a plethora of species, some of them occupying a same ecological niche. As much as conspecifics could be competitors, or direct or indirect cooperators, so could be individuals from another species with similar ecological preferences, for instance with regards to diet. As such, co-occurrence of species from the same guild might contribute into shaping animal cognition. On the one side, the presence of sympatric species with overlapping diet could contribute in reinforcing selective pressures for advanced cognition because (i, competition) species would compete for food access (i.e. Red Queen paradigm, [@van1973new]). Co-occurrence would increase the environmental complexity due to impoverishment of food and addition of noise to the spatio-temporal availability signal because of unforeseen depletion. (ii, cooperation/exploitation) sympatric species presence cues could also represent additional information to process to infer resource location and availability. Logically then, species with higher brain size would be the most evolutionary successful, thus “booming” and intensively diversifying as for hominins [@melchionna2020macroevolutionary]. On the other side, (i) the increase in environmental complexity could be such that advanced cognition is no longer adaptive (see for instance [@grove2013evolution] and [@robirainreview] for the limit of the adaptiveness of spatial and temporal cognition respectively), or (ii) the additional cues provided by other species presence would not add, but replace, and perhaps be more easily interpreted than, environmental cues of food availability. In this latter case, selective pressure on cognitive abilities would be relaxed, inducing a decrease in brain size in sympatric species compared to lonely species. Following this rationale, species living in co-occurrence or not with other species would not face the same evolutionary fate despite initially similar environmental conditions. Thus, the evolutionary success of specific lineages would end up unrelated to brain size. In this study, we therefore aimed to test how species co-occurrence contributed to shaping the evolutionary history of the encephalisation of the whole, or part of, the brain, and whether this induced a "boom" or a "brake" to cognitive abilities and associated evolutionary success by focusing on frugivorous primates as a study example.

@@

@@

|   Recent tools have been developed to infer the effect of species interactions on trait evolution, either by modelling trait divergence in co-occurring species from a same guild (e.g. dietary guild; Matching Competition: MC models) or considering that the evolutionary rate depends on the number of lineages within the guild (density dependence; linear: DD$_{lin}$ or exponential: DD$_{exp}$; @drury2016estimating). After reconstructing primate biogeography history when considering `r length(areaName)` biogeographic areas following [@kamilar2009environmental] based on `r minRange` primate species (@matzke2013probabilistic; @matzke2016stochastic; Figure \@ref(fig:figmap)) as well as primate diet evolution based on `r minFruit + minLeaf` to `r maxFruit + maxLeaf` species (discrete trait: frugivory vs. folivory; @bollback2006simmap. The classification varied depending on how frugivory/folivory was assessed, see [Dietary guild]), we calculated the likelihoods of models considering the role of species interactions (including competitive scenarios) in the evolution of either the whole brain (using the encephalic quotient, EQ, as a proxy for `r minEQ` to `r maxEQ` frugivorous), or the relative size of specific brain areas associated with foraging-related information perception, processing or retention (Figure. \@ref(fig:figbrain); comprising `r minAllAreas` to `r max(maxNeocortex,maxHippocampus,maxCerebellum,maxStriatum,maxMOB)` frugivorous species). The use of specific region size relatively to the body mass (see @deaner2000comparative for further consideration of scaling methods), and not raw size, rather depicts the evolutionary evolution of cognitive abilities in terms of allocation rather than abilities per se (although it is vividly discussed whether raw measures are anything informative on "abilities" too @logan2018beyond). We also estimated the likelihoods of simpler models assuming no effect of species interactions, like the Brownian Motion (BM), the Ornstein-Uhlenbeck process (OU) assuming that traits are constrained around on optimal value (e.g. stabilizing selection; see @blomberg2020beyond for a review on these approaches) or the Early-Burst model (EB, @blomberg2003testing), this latter allowing to check for a time-dependence of the evolutionary rate, hence emphasizing that, if any, the density effect is not an artefact due to time dependence. Support for each model was evaluated using an information-theoretic framework [@burnham2002model] based on the weight of Akaike Information Criterion corrected for small samples (AICc) when considering all six models (MC, DD$_{lin}$, DD$_{exp}$, BM, OU, EB, see [Models of trait evolution: does interspecific interactions shape brain size evolution?]). Non-competitive models were the most likely in describing the evolutionary history of the EQ, the Neocortex and the Cerebellum (Figure \@ref(fig:figbrain) and \@ref(fig:figresultsevolution)), two areas specifically involved in movement and/or immediate information processing [@wiltgen2004new; @koziol2014consensus; @sokolov2017cerebellum] but also in memory consolidation for the Neocortex [@wiltgen2004new]. By contrast, competitive models were most supported in areas involved in sensory abilities (the main olfactory bulb, MOB), short-term working memory and long-term spatio-temporal information retention (Hippocampus, @burgess2002human), and the Striatum an area that is involved in information processing during social interaction (i.e. social reward assessment; @izuma2008processing) (Figure \@ref(fig:figbrain) and \@ref(fig:figresultsevolution)). When density-dependent models were the best fit, the rate (*r*, Figure \@ref(fig:figresultsevolution)) suggested an acceleration of the evolutionary tempo together with increased lineage diversity for the Hippocampus and Striatum, but a slow down for the MOB.

@@

|   Next, to understand the directionality of the selection gradient shaped by co-occurrence (i.e. selection for “bigger” or “smaller” brain if the more species), we fitted phylogenetic regressions (see [Phylogenetic regressions] a)). For these linear regressions, the predicted variable was the relative brain size values of the different areas. We considered the average surface of the frugivorous species range that was overlapped by other sympatric frugivorous species, as well as the number of such sympatric frugivorous species across their entire range as covariates. On average ($\pm$ SE), the considered primate species had `r meanNspeciesCoocc` in their distribution range ($\pm$ `r seNspeciesCoocc`). That ranged from `r minNspeciesCoocc` other species (*`r whichSpeciesMinNCoocc`*), to `r maxNspeciesCoocc` species (*`r whichSpeciesMaxNCoocc`*). On average ($\pm$ SE), the considered primate species had `r meanOverlap*100`% of their range overlapping with other species ($\pm$ `r seOverlap*100`). That ranged from `r minOverlapSpecies*100`% of overlap (*`r whichSpeciesMinOverlap`*), to `r maxOverlapSpecies*100`% of overlap (*`r whichSpeciesMaxOverlap`*). The number of sympatric species never influenced significantly the relative size of the brain or other specific areas (Table \@ref(tab:tableRegGradient)). The percentage of range shared on average with other species was coherently assessed as significantly correlated, or as a trend, with the relative size of areas which evolutionary history was better described with competitive models: the Hippocampus, the MOB and the Striatum (Hippocampus: $t$`r testGradientHippocampus`;  MOB: $t$`r testGradientMOB`; Striatum: $t$`r testGradientStriatum`). The correlations were all negative (Hippocampus: `r estimateGradientHippocampus`;  MOB: `r estimateGradientMOB`; Striatum: `r estimateGradientStriatum`), which means that higher overlap rhymes with lower relative size, insensitive to data and the phylogeny variability (Table \@ref(tab:tabledfsensitivity)). Thus, it suggests that sympatric species are subject to less stringent selection on advanced cognitive abilities.

@@

@@

|   Finally, we investigated how brain evolution was related to evolutionary success, assumed proportional to diversification rate (i.e. speciation minus extinction rate), by using birth-death models of species diversification [@maliet2019model]. Overall, diversification, estimated based on molecular phylogeny without fossil records, increased over time (Figure \@ref(fig:figdiversificationTime), particularly in the early and late Miocene, around `r dateFirstRupt[2]` (CI95% = `r paste("[",dateFirstRupt[3],",",dateFirstRupt[1],"]", sep="")`) and `r dateSecondRupt[2]` (CI95% = `r paste("[",dateSecondRupt[3],",",dateSecondRupt[1],"]", sep="")`) Myr ago (Figure \@ref(fig:figdiversificationTime)). Visual inspection clearly suggested positive relationship between diversification rate and the size of brain areas (Figure \@ref(fig:figRegressionDiversification)). Yet, accounting for phylogenetic dependence erased such pattern: In fitted Bayesian regressions, the size of brain size was never significantly associated with an increase in diversification rate (Table \@ref(tab:tableRegDiversification); see robustness in Table \@ref(tab:tabledfsensitivity2)).

@@

|   The use of brain size as a proxy for cognition is a central debate with no optimal solution (see grounded criticism from @deaner2000comparative; @healy2007critique; @logan2018beyond). The current flourishment of consortia, allowing for much more detailed and standardized anatomical measurements (e.g. in primates: @milham2018open), or with standardized behaviourally explicit comparisons (e.g. on captive [@many2019establishing] or wild [@janmaat2021using] primates), might alleviate biases stemming from brain size analysis, but this will take time. In the meanwhile, brain size is a proxy much appreciated in practice, because of its easy accessibility for a “large” number of species. Further, biases might be limited by considering measurement variability [@logan2018beyond] or the mosaic structure of the brain [@barton2000mosaic; @decasien2019primate]. We did both. Although it existed a variability in the data (phylogenetic and on traits), results were robust. In addition, we saw that the evolutionary history between specific brain regions did not equally depend on the number of lineages living in sympatry. The effect of between-species interaction was indeed only evidenced for specific areas, more particularly those involved in immediate information processing based on senses (Main Olfactory Bulb, MOB), in a working memory or in a long-term memory of spatio-temporal information (Hippocampus) and in processing social cues (Striatum). These areas thus imply individual-based and social-based information processing, pinpointing that the two components might be under selection in primates. This supports the general discussion on the importance of social vs. ecological factors to explain primate cognition evolution and diversity [@decasien2017primate; @powell2017re]. Using a modelling approach including metabolic, life-history and game theories, @gonzalez2018inference emphasized that ecological challenges were preponderant (equating around 60% of challenges faced) to explain current human brain size, which then was also substantially promoted by the occurrence of social challenges (equating around 30% of challenges faced). Here, we highlighted that the cognitive function allowing processing sociological or ecological cues are both affected by species sympatry.

@@

<!-- From small to long range ( < few km) primates might rely on sensory abilities to decide where to go [@dominy2001sensory]. They are known as visual mammals. In particular, frugivorous species highly benefit from a trichromatic vision that enables them to distinguish fruits within the dense foliage [@osorio1996colour]. Yet, in tropical forests, visual range is easily impeded, being barely greater than a hundred of meters. On the contrary olfactory cues are of longer distance.  -->

>>>>

---
title: \Huge \textbf{Species sympatry shapes brain size evolution in Primates}
# date: "`r format(Sys.time(), '%B %d, %Y')`"

#I follow: https://stackoverflow.com/questions/52918716/authors-and-affiliations-in-the-yaml-of-rmarkdown for author display
author: 
  - name: \textit{Benjamin Robira}
    email: benjamin.robira@normalesup.org
    institute: [CEFE, MH]
    correspondence: true
  - name: \textit{Benoit Perez-Lamarque}
    email: benoit.perez@ens.psl.eu
    institute: [ENS, MNHN]
institute:
  - CEFE: Centre d’Écologie Fonctionnelle et Évolutive, Université de Montpellier & CNRS, Montpellier, France.
  - MH: Anthropologie et Ethnobiologie, Centre National de la Recherche Scientifique/Muséum National d'Histoire Naturelle, University Paris Diderot, Sorbonne Paris Cité, Musée de l'Homme, Paris, France.
  - ENS: Institut de Biologie de l’École Normale Supérieure (IBENS), École Normale Supérieure, CNRS, INSERM, Université PSL, Paris, France.
  - MNHN: Institut de Systématique, Évolution, Biodiversité (ISYEB), Muséum National d’Histoire Naturelle, CNRS, Sorbonne Université, EPHE, Université des Antilles, Paris, France.

@@

bibliography: bibliographyarticlepackage.bib
csl: nature.csl
#biblio-style: apa
always_allow_html: true
link-citations: yes

@@

header-includes:
# - \usepackage[nolists, nomarkers,tablesfirst]{endfloat} # For figures and tables at end
  #- \usepackage{lineno} # For line numbering
  #- \linenumbers # For line numbering
  #- \usepackage{setspace}\doublespacing
  - \usepackage{fontawesome} #for fa symbols
  - \usepackage{tcolorbox}
  - \pagenumbering{gobble} #for no page numbering
  - \setlength{\parskip}{0em} #to remove line gaps
  - \DeclareUnicodeCharacter{2212}{-}
  - \usepackage{caption}
  - \captionsetup[figure]{font=small}
  - \newcommand{\beginsupplement}{
      \setcounter{table}{0}  
      \renewcommand{\thetable}{S\arabic{table}} 
      \setcounter{figure}{0} 
      \renewcommand{\thefigure}{S\arabic{figure}}
    }
---

@@

\newtcolorbox{graybox}{
  colback=lightgray!25,
  colframe=lightgray!25,
  coltext=black,
  boxsep=3pt,
  arc=0pt}

:::: {.graybox data-latex=""}

**Abstract |** In hypotheses related to the evolution of animal intelligence, a particular emphasis was given to the role of intra-species conspecifics. Yet, one species rarely occupies an area alone: Space is a place shared between many species. These latter can be new competitors to outperform or cooperators to decode and maneuver, thereby stimulating cognition. They can also contribute to modifying and depleting the environment such as it overly increases its complexity, thereby alleviating the benefice of cognition. Considering brain size as a proxy for cognition, we used primates as a study example to test for the intertwine between species co-occurrence and cognition. We retraced the evolutionary history of several brain areas considering competitive or non-competitive evolutionary scenarios. We found that competitive evolutionary scenarios best predicted the evolution of some brain areas that were related to interacting with the social or ecological environment, with a decrease of their relative size the higher the sympatry rate. The whole brain size, as well as the size of areas used in more general contexts were not related to the sympatry rate. In any case, the evolutionary success of primate species was unrelated th the relative sizes of brain areas. Overall, this comparative study suggests that species co-occurrence contributes to shaping the selective pressure acting upon animal cognition, yet leaves open the question of the underlying ecological mechanisms at play.

@@

\faTags \hspace{0.01in} **Keywords:** Brain size - Cognition - Competition - Diversification - Intelligence evolution - Primates - Species co-occurrence

<!-- TC:ignore -->
\faInfoCircle  \hspace{0.01in} **Word Count (main text only):** `r words$num_words` \newline
<!-- Character Count: `r words$num_char` -->
<!-- TC:endignore -->

::::

@@

@@

On the road to brain size evolution, generally considered as an equivalent of cognition evolution, mysteries are plenty [@van2006some; @dunbar2017there]. Overall, the evolution of the brain is the consequence of constraints (e.g. energetic limitations, @navarrete2011energetics), but also of socio-ecological drivers promoting cognitive abilities [@gonzalez2018inference]. Primates are pivotal species to study the evolution of cognition given the complexity of their socio-ecological environment and the inevitable implication for retracing human evolutionary history [@byrne2000evolution]. Their cognition can result from interactions of individuals with conspecifics within the social unit (*Social Intelligence Hypothesis*, @byrne1994machiavellian; @dunbar2017there), among generations (*Cultural Intelligence Hypothesis*, @whiten2007evolution; @reader2002social, also known as the *Vygotskian Intelligence Hypothesis* in humans, @herrmann2007humans; @tomasello2009cultural; @wilson1991molecular; @van2011social), between social units (*Napoleonic Intelligence Hypothesis*, @ashton2020interactions), or with the rest of their environment (*Ecological Intelligence Hypothesis*; @clutton1980primates; @milton1981distribution; @rosati2017foraging).

|   A particular emphasis was given to the role of intra-species conspecifics in shaping animals cognitive abilities. Yet, space is a place shared between a plethora of species that can interact directly or indirectly, in particular if sharing some ecological traits, especially diet [@drury2018contrasting]. To settle to new coral reefs, fishes use pops and clicks of other fishes as an honest signal for resourcefulness there [@gordon2019acoustic], mangabeys follow calls from hornbills to locate fruiting trees [@olupot1998fruit], and interactions even happen across kingdoms, with migratory birds interpreting phenological cues as synonymous of insect availability [@mcgrath2009flower].

|   Retracing the evolutionary history of cognitive abilities proves to be challenging because there is still no consensual measurement applicable across all species. Up to now, a raw approximation consists in considering brain size as a proxy for cognitive abilities (but see discussion in @logan2018beyond). Although generally considered as a singular entity, the brain is a mosaic of areas cognitively specialized [@barton2000mosaic]. For instance, the Striatum is stimulated during social interaction [@baez2013role]. Therefore, its size is expected to be positively influenced when contacts with other species increase, for instance in species performing mixed-species groups (e.g. in callhitrichine primates, @heymann2015unlike). Conversely, the evolution of the Main Olfactory Bulb (MOB), allowing processing olfactory cues, or the Hippocampus, home of a spatio-temporal memory [@burgess2002human], should be linked to environmental complexity faced by a species, in particular when foraging. This environmental complexity can increase if space is shared with species with similar dietary habits because it contributes to impoverishing the environment and increasing its unpredictability, thus looping on selective pressure towards cognitive abilities [@grove2013evolution: @robirainreview]. Given high energetic costs of the brain [@raichle2006brain], it is thus expected that sympatry should induce positive selection on the size of brain areas involved in foraging-related information processing, unless the induced environmental complexity starts being too high for cognitive foraging to be adaptive. Given that the Striatum also underpins reward expectation and action, goal-directed behaviour and planning abilities [@johnson2007integrating], the same scenario might apply as well to the Striatum of species that do not live in mixed-groups, which actually constitute the majority of the primate lineage. In between, more generalist areas, such as the Cerebellum [@koziol2014consensus; @sokolov2017cerebellum] and the Neocortex [@wiltgen2004new] underlying movement and/or general information processing and retention, could also be stimulated by the presence of other species movement traces, noises or odors to decode.

|    Recent tools have been developed to infer the effect of species interactions on trait evolution, either by modelling trait divergence in co-occurring species from a same guild (e.g. dietary guild; Matching Competition: MC models) or considering that the evolutionary rate depends on the number of co-occurring lineages within the guild (density dependence; linear: DD$_{lin}$ or exponential: DD$_{exp}$; @drury2016estimating). After reconstructing primate biogeography history (N$_{species}$ = `r minRange`, @matzke2013probabilistic; @matzke2016stochastic) when considering `r length(areaName)` biogeographic areas (Figure \@ref(fig:figmap); @kamilar2009environmental) and diet evolution (N$_species$ = `r minFruit + minLeaf` to `r maxFruit + maxLeaf`; @bollback2006simmap), we calculated the likelihoods of models considering the role of species interactions (including competitive scenarios) in the evolution of either the whole brain (using the encephalic quotient, EQ, as a proxy, N$_{species}$ = `r minEQ` to `r maxEQ`), or the relative size of specific brain areas associated with foraging-related information perception, processing or retention (Figure. \@ref(fig:figbrain); N$_{species}$ = `r minAllAreas` to `r max(maxNeocortex,maxHippocampus,maxCerebellum,maxStriatum,maxMOB)`). The use of specific region size relatively to the body mass and not raw size depicts the evolution of cognitive abilities in terms of allocation rather than abilities per se (but see discussion in @deaner2000comparative). We also estimated the likelihoods of simpler models assuming no effect of species interactions, like the Brownian Motion (BM), the Ornstein-Uhlenbeck process (OU) considering that traits are constrained around on optimal value (e.g. stabilizing selection; see @blomberg2020beyond for a review) or the Early-Burst model (EB, @blomberg2003testing), this latter allowing to check for a time-dependence of the evolutionary rate, hence emphasizing that, if any, the density effect is not an artefact due to time-dependence. Support for each model was evaluated using an information-theoretic framework [@burnham2002model] based on the weights of the Akaike Information Criterions corrected for small samples (AICc) when considering all six models (MC, DD$_{lin}$, DD$_{exp}$, BM, OU, EB, see [Models of trait evolution: does interspecific interactions shape brain size evolution?]).

@@

|   Non-competitive models were the most likely in describing the evolutionary history of the EQ, the Neocortex and the Cerebellum (Figure \@ref(fig:figbrain) and \@ref(fig:figresultsevolution)), two areas specifically involved in movement and/or general information processing [@wiltgen2004new; @koziol2014consensus; @sokolov2017cerebellum] but also in memory consolidation for the Neocortex [@wiltgen2004new]. By contrast, competitive models were most supported in areas involved in short-term working memory and long-term spatio-temporal information retention (Hippocampus, @burgess2002human) and in planning, reward assessment, and information processing during social interaction (Striatum, @baez2013role; @johnson2007integrating), while they tended to be so for one involved in sensory abilities (the main olfactory bulb, MOB; Figure \@ref(fig:figbrain) and \@ref(fig:figresultsevolution)). When density-dependent models were the best fit, the rate (*r*, Figure \@ref(fig:figresultsevolution)) suggested an acceleration of the evolutionary tempo together with increased lineage diversity for the Hippocampus and the Striatum, but a slowdown for the MOB.

@@

|   Next, to understand the directionality of the selection gradient shaped by co-occurrence (i.e. selection for “bigger” or “smaller” brain if the more species), we fitted phylogenetic regressions (see [Phylogenetic regressions] a)). For these linear regressions, the predicted variable was the relative brain size values of the different areas. We considered the average surface of the frugivorous species range that was overlapped by other sympatric frugivorous species, as well as the number of such sympatric frugivorous species across their entire distribution range as covariates. On average ($\pm$ SE), the considered primate species had `r meanOverlap*100`% of their range overlapping with other species ($\pm$ `r seOverlap*100`). That ranged from `r minOverlapSpecies*100`% of overlap (*`r whichSpeciesMinOverlap`*), to `r maxOverlapSpecies*100`% of overlap (*`r whichSpeciesMaxOverlap`*). In term of distribution range, the considered primate species co-occurred on average with `r meanNspeciesCoocc` other primate species ($\pm$ `r seNspeciesCoocc`), ranging from `r minNspeciesCoocc` other species (*`r whichSpeciesMinNCoocc`*), to `r maxNspeciesCoocc` species (*`r whichSpeciesMaxNCoocc`*). The number of sympatric species never influenced significantly the relative size of the brain or other specific areas  (Table \@ref(tab:tableRegGradient)). Conversely, we found that the percentage of range shared with other species correlated with the relative size of areas for which the evolutionary history was better described with competitive models: The Hippocampus, the MOB and the Striatum (Hippocampus: $t$`r testGradientHippocampus`;  MOB: $t$`r testGradientMOB`; Striatum: $t$`r testGradientStriatum`). The correlations were all negative (Hippocampus: `r estimateGradientHippocampus`;  MOB: `r estimateGradientMOB`; Striatum: `r estimateGradientStriatum`), which means that higher species overlap associates with lower relative size, insensitive to data and phylogeny variability (Table \@ref(tab:tabledfsensitivity)). Thus, it suggests that sympatric species are either subject to a positive selection for smaller brain, or to a less stringent selection on advanced cognitive abilities.

@@

# bp.resp <- breakpoints(Diversification ~ Time, data=aggregatedSpeciationTime.mean)
# summary(bp.resp)
# ## the BIC 
# plot(bp.resp)
# breakpoints(bp.resp)

#Suggest that 4 is better from stats, visually 2

@@

|   Finally, we investigated the evolutionary consequences of brain evolution by evaluating whether brain sizes correlated with species diversification rates (i.e. speciation minus extinction rates), by using lineage-specific birth-death models of species diversification [@maliet2019model; @maliet2020fast]. Overall, species diversification, estimated based on molecular phylogeny increased over time (Figure \@ref(fig:figdiversificationTime)), particularly in the early and late Miocene, around `r dateFirstRupt[2]` (CI95% = `r paste("[",dateFirstRupt[3],",",dateFirstRupt[1],"]", sep="")`) and `r dateSecondRupt[2]` (CI95% = `r paste("[",dateSecondRupt[3],",",dateSecondRupt[1],"]", sep="")`) Myr ago (Figure \@ref(fig:figdiversificationTime)). When accounting for phylogenetic dependence, no significant relationship between diversification rate and the relative size of brain areas was found (Table \@ref(tab:tableRegDiversification); see robustness in Table \@ref(tab:tabledfsensitivity2)).

@@

@@

The bigger is not necessarily the better. The size of the brain is subject to a compromise between the energy it incurs, and the increase of fitness it allows. This is clearly emphasized by the fact that the biggest areas, the Cerebellum and the Neocortex, as well as the whole brain (EQ) were best described by the Orstein-Uhlenbeck process, what might suggest a stabilisation towards an optimal size as a result of an equilibrium between costs and benefices. As the brain area is regionally specialised [@barton2000mosaic], it was also expected that brain regions could be under different selective pressures. For instance, the Lemuriformes, that are known to prioritize smell compared to other primate species, indeed have the largest relative MOB size (i.e. pondered by body size) in our data (Lemuriformes: mean $\pm$ SE = `r meanseMOBlemu[1]` $\pm$ `r meanseMOBlemu[2]`, other: `r meanseMOBother[1]` $\pm$ `r meanseMOBother[2]`, \@ref(fig:figbrain)). Similarly, Platyrrhini, and Callitrichine in particular, are known to form poly-specific associations [@heymann2000behavioural] and indeed show the highest relative size of the Striatum in our data (Platyrrhini: mean $\pm$ SE = `r meanseStriatumplaty[1]` $\pm$ `r meanseStriatumplaty[2]`, other: `r meanseStriatumother[1]` $\pm$ `r meanseStriatumother[2]`, \@ref(fig:figbrain)). Here, we show that sympatry is one factor that affects the selective regime under which each brain region evolves: Although the brain as a whole was insensitive to species co-occurrence, this latter nonetheless induced a change in the relative size of the Hippocampus, and the Striatum, and potentially in the MOB. These areas imply individual-based and social-based information processing, pinpointing that the two components might be under selection in primates. This supports the general discussion on the importance of social *vs* ecological factors to explain primate cognition evolution and diversity [@decasien2017primate; @powell2017re].

|   The relative size of these brain areas were negatively associated with one index of sympatry intensity, the average overlap of a species distribution range with that of other species. Because the ratio of the other brain areas is insensitive to sympatry intensity, it is unlikely that this results from a by-product of a selection on body mass and might thus be a consequence of species direct or indirect interactions. In general, competition is the first-thought mechanism to describe community structures [e.g. @rocha2015role]. In primates, it is rare to observe direct competition between species, which should translate into site exclusion. In fact, inter-species site exclusion in this taxa has only been observed in gibbons [@suwanvecho2012interspecific]. For species with overlapping diet, indirect competition for resource might happen. It has been shown that individuals tend to use social or environmental cues depending on their reliability [@rafacz2003environmental; @dunlap2016foraging]. The utilization of these cues should promote cognition until reaching a certain range of environmental complexity [@grove2013evolution; @robirainreview]. Our results could thus suggest that species sympatry may trigger an important decrease in cues reliability, for instance due to higher and unforeseen depletion rate. That would explain a relaxed positive selection for advance cognitive abilities, and conversely, given the cost of brain maintenance [@raichle2006brain], would induce a negative selection towards areas involved in foraging related activities like the Hippocampus, the MOB and the Striatum.
